The fundamental shape of dynamic experiences is preserved in episodic memory

### 2018-06-26 main.tex
checking in

To study the temporal and semantic aspects of memory for naturalistic experiences, we analyzed an open dataset in which participants view and verbally recall an episode of the BBC series \textit{Sherlock}  \citep{ChenEtal17}. We developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17} to capture the temporal evolution of rich semantic structure present in naturalistic experiences. We then use this model to characterize the shape and contents of episodic memories for naturalistic experiences. We hypothesized that memory for naturalistic experiences is less about recalling specific details and more about preserving the overall ``shape'' of an experience.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/1_analysis_schematic.pdf}
\caption{\small \textbf{Schematic of the analysis approach.} For each moment of the video, text descriptions were manually generated. Three exemplary time points are displayed here.  Below the video descriptions are text samples from an example participant's verbal recall transcript.  We trained a topic model on the moment-by-moment video description text and transformed participant's recall transcripts using this same model. The bar charts display the resulting topic model weights for the video (in blue) and recall (in orange) for three example topic dimensions.}
\label{fig:schematic}
\end{figure}

\subsubsection{A model to capture the geometric structure of naturalistic stimuli}
We fit a topic model \citep{BleiEtal03} to manually annotated text descriptions of scenes from the episode. The text descriptions contained details of the scene such as the characters, location, and a short summary of the scene (see Fig.\ref{fig:schematic} for example text, see \ref{sec:methods} for analysis details). We then transformed the text descriptions using the (same) topic model, resulting in a scenes (1000) by topics (100) matrix, where each row of the matrix represents a probabalistic mixture of topics discovered in that scene (See Fig.\ref{fig:schematic} for example topic vectors). We expanded this matrix from 1000 to 1976 timepoints by copying the vectors for scenes that spanned multiple TRs. As depicted in Fig.~\ref{fig:model}a, the topics are sparse and change slowly over time. Furthermore, there are clear transitions from one topic `state' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed the timepoint-by-timepoint correlation matrix of the video model (Fig.~\ref{fig:model}c).  This correlation matrix reveals that the model has a strong, block-diagonal structure. Another noteworthy characteristic is that there is very little correlation between blocks (i.e. the off-diagonal values are small), suggesting the representations of each block are unique and highly descriminable.

\subsubsection{Modeling verbal free recall}
After watching the episode, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. The result is a sentences (range: 68 to 294, mean: 131.94) by topics (100) matrix for each participant, where each row represents the estimated mixture of topics for a given window of sentences during recall. An example participant's (#13) model is plotted in Fig.~\ref{fig:model}c. Note that the topics were derived solely from the text descriptions of the video, and so the topic models estimated for recall are dependent on the topics in the episode. This approach effectively projects the participants' verbal recall into the video ``stimulus space'', allowing us to quantify the dynamics of verbally reported memories as well as relate them to their originating experiences.

Next, we investigated the temporal structure of the recall matrices. For each participant, we computed a timepoint-by-timepoint correlation matrix from recall models. An example participant's (\#13) correlation matrix is plotted in Fig.~\ref{fig:model}d). Like the video model, each participant's recall correlation matrix exhibited a block-diagonal structure (Supp Fig X).

\subsubsection{Segmenting the video and recall models into `events'}
One striking feature of the video and recall correlation matrices (see Fig.~\ref{fig:model}) is a strong, block structure along the diagonal of the matrices.  We hypothesized that this structure might arise from transient stability in the content of the video, and the subsequent language used to describe each event of the video. If true, we reasoned that it would be possible to infer the specific event a participant is describing by comparing the (average) topic model for a particular recall ``event'' with the (average) topic vector for each video event. In other words, the video event that is most similar to a given recall event is the event that the participant is most likely describing.  To test this idea, it is first necessary to systematically parse the matrices into events.  We leveraged a hidden Markov model (~\citep{BaldEtal17}) to estimate $k$ events (or states) from a (multivariate) time series. The algorithm determined 34 events for the movie model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see ~\ref{sec:methods} for details on choosing an optimal $k$ value).  The events discovered for a representative participant (\#13) are highlighted as yellow rectangles outlining blocks along the diagonal of the participant's correlation matrix (Fig.~\ref{fig:model}d). Next, we created a video event model by averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}e).  We performed the same procedure for the recall matrices (see Fig.~\ref{fig:model}f for example). Then, we computed the correlation between these two event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (see Fig/~\ref{fig:model}g). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video events the recall events most likely referred to, for each recall event, we found the index of the video event with the highest correlation (i.e. the argmax).  This is depicted in Fig.~\ref{fig:model}g as the cell highlighted with the yellow box for each column in the video-recall event correlation matrix. Notably, our algorithm suggests that the example participant recalled the events largely in order (occasionally skipping over an event).

Next, we computed a group-averaged recall event model and video-recall event correlation matrix.  To do this, for each participant (and each recall event), we grouped the recall event vectors (across all participants) by the video event that they were most highly correlated to. We then averaged the recall event vectors within group. This yielded an average recall event vector for all but one (of 34) event (since no participant remembered this event according to our model). Lastly, we computed an average recall event (34) by video event (34) correlation matrix, and highlighted the highest correlation in the column with a black box (Fig.~\ref{fig:model}h). Notably, this matrix displayed high correlation values along the diagonal and low correlations in the off-diagonal cells. This suggests that on average, participants were able to recapitulate the events in the episode and our approach captured this behavior.
% NOTE: ^ possibly need a null model for the group analysis.  I think it could be to get a result that looks like this even if we scrambled the text. My concern is that I'm grouping the recall events by the video event they were most similar to, and then averaging, which may give rise to a smooth looking plot without any real information there...

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{figs/2_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} For all plots, darker colors indicate greater values and the range of each plot is 0-1.  A). A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint. Each columns represents a different topic. B). A timepoints (1976) by topics (100) matrix representing participant \#13's recall. C). A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The white boxes represent `events' recovered by a hidden Markov model. D). A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. E). An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  F). An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. G. A recall events (27) by video events (34) correlation matrix for participant \#13. The yellow squares identify the video event with the highest correlation to a given recall event. F. A group averaged recall events (34) by video events (34) correlation matrix.  The yellow squares identify the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{figure}

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/3_list_learning.pdf}
\caption{\small \textbf{Naturalistic extensions of classic memory analyses.} A). The probability of first recall as a function of the serial position of the event during encoding. B). Given recall of event i, the probability that the next recalled item will be from serial position i +/- lag. C). Proportion recalled as a function of serial position. All error bars are the standard error of the mean derived from a bootstrap resampling procedure.}
\label{fig:list-learning}
\end{figure}

\subsubsection{Naturalistic extensions of classic memory analyses}
Just like in a traditional `free-recall' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching approach described above affords us the ability to analyze memory in the same way (where the recalled words are analogous to recalled events). In our first set of analyses, we sought to characterize memory performance/dynamics by extending classic analyses originally designed for list-learning experiments to more naturalistic settings.

>>>>

To study the temporal and semantic aspects of memory for naturalistic experiences, we analyzed an open dataset in which participants view and verbally recall an episode of the BBC series \textit{Sherlock}  \citep{ChenEtal17}. We developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17} to capture the temporal evolution of rich semantic structure present in naturalistic experiences. We then use this model to characterize verbally recalled episodic memories for naturalistic experiences. We hypothesized that memories for naturalistic experiences are 1) segmented and centered around particular events and 2) less about recalling specific details and more about preserving the overall ``shape'' of an experience.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/1_analysis_schematic.pdf}
\caption{\small \textbf{Schematic of the analysis approach.} For each scene in the video, text descriptions were generated by hand. Three exemplary time points are displayed here.  Below the video descriptions are text samples from an example participant's verbal recall transcript.  We fit a topic model to the moment-by-moment video text descriptions and transformed participant's verbal recall transcripts using this same model. The bar charts display the resulting topic model weights for the video (in blue) and recall (in orange) for three example topic dimensions.}
\label{fig:schematic}
\end{figure}

\subsubsection{A model to capture the temporal and semantic structure of naturalistic stimuli}
We fit a topic model \citep{BleiEtal03} to (manually annotated) text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters, location, and a short summary of the scene (see Fig.\ref{fig:schematic} for example text, see \ref{sec:methods} for analysis details). We then transformed the text descriptions using the (same) topic model, resulting in a scenes (1976) by topics (100) matrix, where each row of the matrix represents a probabalistic mixture of topics discovered in that scene (See Fig.\ref{fig:schematic} for example topic vectors). As depicted in Fig.~\ref{fig:model}a, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic `state' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoint (1976) by timepoint (1976) correlation matrix of the video model (Fig.~\ref{fig:model}c).  This correlation matrix reveals that the model has a strong, block-diagonal structure. Another noteworthy characteristic is that there is very little correlation between blocks (i.e. the off-diagonal values are small), suggesting the representations of each block are unique and highly descriminable.

\subsubsection{Modeling verbal recall}
After watching the video, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. The result was a sentences (range: 68 to 294, mean: 131.94) by topics (100) matrix for each participant, where each row represents a probabalistic mixture of topics for a given window of sentences during recall. An example participant's (#13) recall model is plotted in Fig.~\ref{fig:model}c. Like the video model, topic vectors were sparse and changed gradually.  Note that the topics were derived solely from the text descriptions of the video, and so the topic models estimated for recall are dependent on the topics present in the video. This approach effectively projects the participants' verbal recall into the video's `topic space'', allowing us to directly compare them to the video model.

Next, we investigated the temporal structure of the recall matrices. For each participant, we computed a timepoint (68 to 294) by timepoint (68 to 294) correlation matrix from the recall models. An example participant's (\#13) correlation matrix is plotted in Fig.~\ref{fig:model}d). Like the video model, each participant's recall correlation matrix exhibited a strong block-diagonal structure (Supp Fig X). Notably, this suggests that participants recounted the video in discriminable segments, potentially indexing the recall of specific events from the video.

\subsubsection{Segmenting the video and recall models into `events'}
As described above, a striking feature of the video and recall correlation matrices (see Fig.~\ref{fig:model}) is a strong, block structure along the diagonal of the matrices.  We hypothesized that this structure might arise from transient stability in the contents of the video (i.e. ``events''), as well as the subsequent language used to recount each event of the video. If true, we reasoned that it would be possible to identify the specific event a participant is describing by comparing the topic vector for a particular recall ``event'' with the topic vector for each video event. In other words, the video event that is most similar (i.e. correlated) to a particular recall event is the event that the participant is most likely describing.

To test this idea, first we segmented the video and recall models in $k$ events (i.e. states) using a hidden Markov model (~\citep{BaldEtal17}). The algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see ~\ref{sec:methods} for details on choosing an optimal $k$ value).  The events discovered for the model and a representative participant (\#13) are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}b,d).

Next, we created a video event model by averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}e).  We performed the same procedure for the recall matrices (see Fig.~\ref{fig:model}f for example). Then, we computed the correlation between video and recall event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (Fig/~\ref{fig:model}g). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video event a particular recall event was most likely describing, we computed the index of the video event with the highest correlation to the recall event (i.e. the argmax).  This is depicted in Fig.~\ref{fig:model}g as the cells highlighted with a yellow border. Interestingly, our algorithm suggests that the example participant recalled the events largely in order, occasionally skipping a video event.

Then, we computed a group-averaged recall event model and video-recall event correlation matrix. For each participant (and each recall event), we sorted the recall event vectors (across all participants) according to the video event that they were most highly correlated to. We then averaged the recall event vectors within each group. This yielded an average recall event vector for all but one (of 34) video events, since no participant remembered one of the events according to our model. Lastly, we computed an average recall event (34) by video event (34) correlation matrix, and highlighted the cell with the highest correlation value with a yellow border (Fig.~\ref{fig:model}h). Notably, this matrix displayed high correlation values along the diagonal and low correlations in the off-diagonal cells. This suggests that on average, participants were able to recapitulate the events in the episode in a specific, highly descriminable way.
% NOTE: ^ possibly need a null model for the group analysis.  I think it could be to get a result that looks like this even if we scrambled the text. My concern is that I'm grouping the recall events by the video event they were most similar to, and then averaging, which may give rise to a smooth looking plot without any real information there...

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{figs/2_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} A depiction of our analysis pipeline. For all plots, darker colors indicate greater values and the range of each plot is 0-1.  A). A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint (i.e. topic weights). Each column represents a different topic. B). A timepoints (1976) by topics (100) matrix representing participant \#13's recall. C). A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The white boxes represent `events' recovered by a hidden Markov model. D). A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. E). An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  F). An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. G. A recall events (27) by video events (34) correlation matrix for participant \#13. The cells with a yellow border identify the video event with the highest correlation to a given recall event. F. A group averaged recall events (34) by video events (34) correlation matrix.  The cells with yellow borders are the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{figure}

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/3_list_learning.pdf}
\caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} A). The probability of first recall as a function of the serial position of the event during encoding. B). A lag-conditional response probability curve. Given recall of event i, the probability that the next recalled item will be from serial position i +/- lag. C). Proportion of events recalled as a function of serial position. All error bars are the standard error of the mean derived from a bootstrap resampling procedure.}
\label{fig:list-learning}
\end{figure}

\subsubsection{Naturalistic extensions of classic list-learning memory analyses}
Just like in a traditional `free-recall' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. The recalled events can be treated as ``items'' analogous to words recalled in a list-learning study. In our first set of analyses, we sought to characterize memory performance/dynamics by extending classic analyses originally designed for list-learning experiments to more naturalistic settings.
